Variable Importance for Oblique Learning Ensembles of Trees (VIOLET)

### 2022-03-07 index.Rmd
swap to berd app

---
title: "Preliminary results: Accelerating ORSF pilot study"
description: |
  Partial model fitting improves computation time and prognostic accuracy for the ORSF.
author:
  - name: Byron C. Jaeger
affiliation: Wake Forest School of Medicine
date: "`r Sys.Date()`"
output:
  distill::distill_article:
  toc: true
toc_depth: 2
code_folding: true
---

@@

rspec <- round_spec() |> 
  round_using_magnitude(digits = c(3, 2, 2, 1),
                        breaks = c(1, 10, 100, Inf))

@@

tar_load(names = data_results)

@@

# Introduction

This document shows the full preliminary results that I have included in a response to reviewers for my submission to the Wake Forest Center for Biomedical Informatics Pilot Award. In my response, I present a table that aggregates results over three datasets. In this document, all three datasets are described and the results for each dataset are tabulated.

# Datasets

## Mayo Clinic Primary Biliary Cholangitis Data, N = 276

Primary sclerosing cholangitis is an autoimmune disease leading to destruction of the small bile ducts in the liver. Progression is slow but inexhortable, eventually leading to cirrhosis and liver decompensation. The condition has been recognised since at least 1851 and was named "primary biliary cirrhosis" in 1949. Because cirrhosis is a feature only of advanced disease, a change of its name to "primary biliary cholangitis" was proposed by patient advocacy groups in 2014.

This data is from the Mayo Clinic trial in PBC conducted between 1974 and 1984. A total of 424 PBC patients, referred to Mayo Clinic during that ten-year interval, met eligibility criteria for the randomized placebo controlled trial of the drug D-penicillamine. The first 312 cases in the data set participated in the randomized trial and contain largely complete data. The additional 112 cases did not participate in the clinical trial, but consented to have basic measurements recorded and to be followed for survival. Six of those cases were lost to follow-up shortly after diagnosis, so the data here are on an additional 106 cases as well as the 312 randomized participants. The `pbc_orsf` data includes 276 participants who had no missing data for the study covariates. The first 10 rows of the data are printed here:

```{r}
as_tibble(pbc_orsf)
```

## Rotterdam Breast Cancer Data, N = 2,982

The `rotterdam` data set includes 2982 primary breast cancers patients whose records were included in the Rotterdam tumor bank. These data sets are used in a paper by Royston and Altman that is referenced below. The first 10 rows of the data are printed here:

```{r}
as_tibble(survival::rotterdam)
```

## Simulated data with interactions, N = 5,000

A simulated dataset with normally distributed predictor variables was generated. Predictor variables $z_1$, $\dots$, $z_{10}$ were negatively associated with the outcome, and $x_1$, $\dots$, $x_{10}$ were positively associated with the outcome. Predictor variables $g_1$, $\dots$, $g_5$ were unrelated to the outcome, but each of the $g_i$ variables had an interaction effect with $x_i$ (i.e., $g_1$ interacted with $x_1$, $g_2$ interacted with $x_2$, and so on, with $g_5$ interacting with $x_5$). An example of these simulated data are presented below.

@@

as_tibble(sim_surv(n=10))

@@

# Methods

I compare four different modeling techniques:

1. `aorsf` - the software I propose to develop further.

1. `obliqueRSF` - the software I previously developed for ORSFs.

1. `randomForestSRC` - A standard package for random forests.

1. `ranger` - Another standard package for random forests.

These techniques are compared based on their discrimination (C-statistic) and scaled Brier score, which are computed using the `riskRegression` package, and specifically `riskRegression::Score()`. Computational time required for model fitting and model predictions are also presented. Results are averaged over 25 runs of Monte-Carlo Cross-validation. In each run of Monte-Carlo Cross-validation, the current data are separated into a training and testing set, all modeling techniques are applied to the training set, and model predictions are evaluated in the testing set.

# Results

Overall, `aorsf` and `obliqueRSF` had the highest prognostic accuracy, but `aorsf` is __much__ faster. In addition, while `obliqueRSF` was more accurate overall than `ranger` and `randomForestSRC`, it was not as accurate as `aorsf`!

@@

results_overall <- data_results |> 
  group_by(model) |>
  summarize(
    across(
      .cols = c(cstat, IPA, time_fit, time_prd),
      .fns = list(mean = mean, sd = sd)
    )
  ) |> 
  mutate(data = 'Overall')

data_tbl <- data_results |> 
  group_by(model, data) |>
  summarize(
    across(
      .cols = c(cstat, IPA, time_fit, time_prd),
      .fns = list(mean = mean, sd = sd)
    )
  ) |> 
  bind_rows(results_overall) |> 
  group_by(data) |> 
  mutate(
    across(.cols = starts_with("time"), .fns = as.numeric),
    time_fit_ratio = time_fit_mean / time_fit_mean[model == 'aorsf'],
    time_prd_ratio = time_prd_mean / time_prd_mean[model == 'aorsf']
  ) |> 
  transmute(
    model, 
    data = recode(
      data,
      'pbc_orsf' = 'Mayo Clinic Primary Biliary Cholangitis Data, N = 276',
      'rotterdam' = 'Rotterdam Breast Cancer Data, N = 2,982',
      'sim' = 'Simulated data with interactions, N = 5,000'
    ),
    cstat = table_glue("{cstat_mean} ({cstat_sd})"),
    ipa = table_glue("{IPA_mean} ({IPA_sd})"),
    time_fit_mean = format(round(time_fit_mean, 4), nsmall = 4),
    time_fit_ratio = table_value(time_fit_ratio),
    time_prd_mean = format(round(time_prd_mean, 4), nsmall = 4),
    time_prd_ratio = table_value(time_prd_ratio)
  )

gt(data_tbl, rowname_col = 'model', groupname_col = 'data') |> 
  cols_label(cstat = 'C-Statistic (SD)',
             ipa = 'Scaled Brier Score (SD)',
             time_fit_mean = 'Mean, seconds',
             time_fit_ratio = 'Ratio',
             time_prd_mean = 'Mean, seconds',
             time_prd_ratio = 'Ratio') |> 
  cols_align(align = 'center') |> 
  tab_spanner(label = 'Model prediction accuracy',
              columns = c('cstat', 'ipa')) |> 
  tab_spanner(label = 'Model fit time',
              columns = c('time_fit_mean', 'time_fit_ratio')) |> 
  tab_spanner(label = 'Risk prediction time',
              columns = c('time_prd_mean', 'time_prd_ratio'))

@@

The novelty of `aorsf` is its use of partially trained models in non-terminal nodes of decision trees. While `obliqueRSF` develops fully trained models, which require iteration until a convergence criteria is met, `aorsf` just completes one iteration and moves along. Initially I thought this computational shortcut would hurt the accuracy of the model, but the reality is that using partially trained models appears to improve the accuracy of the ORSF by increasing the diversity of trees in the ensemble.

>>>>

---
title: "Variable Importance for Oblique Learning Ensembles of Trees (VIOLET)"
description: |
  Preliminary results for BERD pilot application
author:
  - name: Byron C. Jaeger
affiliation: Wake Forest School of Medicine
date: "`r Sys.Date()`"
output:
  distill::distill_article:
  toc: true
toc_depth: 2
code_folding: true
---

@@

rspec <- round_spec() |> 
  round_using_magnitude(digits = c(3, 2, 1, 1),
                        breaks = c(1, 10, 100, Inf))

@@

tar_load(names = benchmark_vi_smry)

@@

## Introduction

This document shows the full preliminary results that I have included in my application for the Biostatistics, Epidemiology, and Research Design (BERD) Program's Methods Pilot Award. The data simulation process is described in full detail, followed by an outline of the proposed simulation study. Last, results are tabulated for each simulation scenario described, and also in aggregate form. The final aggregate table is presented in the BERD application.

## Data Simulation

A simulated data set with five types of predictor variables was generated in each run for the current simulation study. The size of each predictor set was fixed at 25 for our analysis.

- $z$ (irrelevant) variables $z_1$, $\dots$, $z_{25}$ had no relationship with the outcome.

- $x$ (linear effect) variables $x_1$, $\dots$, $x_{25}$ had a linear relationship with the outcome.

- $w$ (non-linear effect) variables $w_1$, $\dots$, $w_{25}$ had a non-linear relationship with the outcome. The non-linear relationship followed a sinusoidal curve with varying period length.

- $g$ (conditional linear) variables $g_1$, $\dots$, $g_{25}$ had a conditional linear relationship with the outcome. In other words, for each $i \in \{1, \ldots, 25\}$,  $g_i$ had no relationship to the outcome, but $g_i * x_i$ was linearly related to the outcome.

- $v$ (combination linear) variables $v_1$, $\dots$, $v_{25}$ had a combination linear relationship with the outcome. In other words, $v_1, v_2, v_3$ each had no effect on the outcome individually, but the latent variable $v_1*c_1* + v_2*c_2 + v_3 * c_3$ was linearly related to the outcome. Coefficients $c_i$ were generated at random uniformly from -1 to -0.5 and 0.5 to 1.

__Mean and standard deviation__ The multivariate normal distribution that generated values for the predictor matrix had a mean of zero and standard deviation of 1 for each variable.

__Correlation__: The correlation among $z$ variables was allowed to be non-zero as was the correlation among $x$ variables and the correlation between $x$ and $z$ variables. For all non-zero correlations, a value was selected at random uniformly between a lower and upper boundary. We simulated five scenarios by varying the boundaries of correlation:

1. lower bound: 0; upper bound: 0
1. lower bound: -0.1; upper bound: 0.1
2. lower bound: -0.2; upper bound: 0.2
1. lower bound: -0.3; upper bound: 0.3
3. lower bound: -0.4; upper bound: 0.4

The correlations were generated symmetrically and were adjusted if needed to create the nearest positive definite covariance matrix using `Matrix::nearPD()`.

__Effect sizes__: A total effect of 3 was distributed evenly across each predictor group. For example, with 25 $x$ predictors, the increase in log-hazard associated with a 1 unit increased in each $x$ variable was $3 / 25 = 0.12$.

__Outcomes__: A censored outcome variable was generated with the `simsurv::simsurv()` using a predictor matrix that included all relevant predictors. After the outcome was generated the interactions, non-linear transformations, and linear combinations of relevant predictors were dropped from the data.

### Demo and validation of simulated data

Some tests of the simulated data are presented below. If you have no skepticism regarding the simulated data, you can safely skip this sub-section. For simplicity, we set the number of predictors to 3 for each predictor group.

__Data generation__: There are no tests for this step, just a glimpse of the simulated set of data.

@@

set.seed(329)

ss_demo <-
  sim_surv(
    n_obs = 5000,
    n_z = 3,
    n_x = 3,
    n_g = 3,
    n_w = 3,
    n_v = 3
  )

glimpse(ss_demo$data)

@@

__Column means__: This test verifies that the mean of each predictor is close to zero, and that roughly 40% of outcomes were censored.

@@

data_sim <- ss_demo$data

data_sim$true_v <- 
  with(
    data_sim,
      v1 * ss_demo$coefs_lc[[1]][1] + 
      v2 * ss_demo$coefs_lc[[1]][2] + 
      v3 * ss_demo$coefs_lc[[1]][3] 
  )

data_means <- apply(data_sim, 2, mean) |> 
  round(3) |> 
  enframe()

# we can see that the mean of each predictor is close to zero, and that roughly 40% of outcomes were censored.
data_means

__Effects detected__: This test verifies that, when the oracle model is used, an effect is only detected for variables that have a relationship to the outcome.

dd <- datadist(data_sim)
options(datadist = dd)

mdl <- cph(
  Surv(time, status) ~ 
    # z variables; no effect should be detected
    z1 + z2 + z3 +
    # x variables; effect should be detected
    # g variables; no effect should be detected
    # interaction between x and g; effect should be detected
    x1*g1 + x2*g2 + x3*g3 + 
    # w variables; effect should be detected and should be nonlinear
    rcs(w1) + rcs(w2) + rcs(w3) + 
    # v variables; no effect should be detected for individual
    # variables v1, v2, and v3, but the linear combination of
    # these variables should have a detectable effect.
    # note: v1 is not included due to co-linearity 
    v2 + v3 + true_v,
  data = data_sim
)

# All expectations outlined above are met!
anova(mdl)

__Effect size estimation for linear effects__: This test verifies that, when the oracle model is used, the estimated effect sizes for variables with linear effects are consistent with the effect sizes used to simulate the data.

effects_expected <- c("x1" = 1, 
                      "x2" = 1, 
                      "x3" = 1,
                      "g1" = 0,
                      "g2" = 0,
                      "g3" = 0,
                      "x1 * g1" = 1,       
                      "x2 * g2" = 1,       
                      "x3 * g3" = 1, 
                      "v2" = 0,
                      "v3" = 0,
                      "true_v" = 1)

effects_observed <- coef(mdl)[names(effects_expected)]

# the differences between observed and expected estimates is small
abs(effects_expected - effects_observed) |> 
  enframe(name = 'predictor',
          value = 'bias')

__Effect size estimation for non-linear effects__: This test verifies that, when the oracle model is used, the estimated effect sizes for variables with non-linear effects are consistent with the effect sizes used to simulate the data.

data_gg <- 
  list(w1 = as.data.frame(Predict(mdl, w1)) |> select(x=w1, yhat),
       w2 = as.data.frame(Predict(mdl, w2)) |> select(x=w2, yhat),
       w3 = as.data.frame(Predict(mdl, w3)) |> select(x=w3, yhat)) |> 
  bind_rows(.id = 'xvar') |> 
  filter(x > -2.5, x < 2.5) |> 
  mutate(xvar = recode(xvar,
                       w1 = "Non-linear variable: w1",
                       w2 = "Non-linear variable: w2",
                       w3 = "Non-linear variable: w3")) |> 
  as_tibble()

# We expect increasing wiggle from left to right.

ggplot(data_gg) + 
  aes(x = x, y = yhat) + 
  geom_line() + 
  facet_wrap(~ xvar) + 
  theme_bw() + 
  labs(x = 'W value', y = 'Estimated log-hazard')

These results verify that the simulated data set has the expected properties 
# Simulation Study

I compare four different techniques to measure variable importance (VI).

1. `Negation importance`: the method I propose to develop further with this pilot application. This is computed using an oblique random survival forest model.

1. `ANOVA importance`: A method introduced by Menze et al. This is computed using an oblique random survival forest model.

1. `SHAP importance`: A method introduced by Lundberg et al. SHAP importance is widely regarded as a gold standard due to its excellent asymptotic properties. This is computed using a boosting model.

1. `Permutation importance`: A method introduced by Leo Breiman. This is computed using a standard random survival forest model.

__Evaluation of variable importance techniques__: VI techniques are compared based on their discrimination (C-statistic) between relevant and irrelevant predictor variables. The relevant predictor variables are those that have a relationship to the outcome ($x$, $w$, $g$, and $v$ variables), and the irrelevant variables are those with no relationship to the outcome, i.e. the $z$ variables. This is equivalent to assigning a binary outcome to each variable: 1 if it is relevant, 0 otherwise, and then computing a C-statistic for discrimination of the relevant versus irrelevant variables using variable importance as a 'prediction' for the variables.

__Details on C-statistic computation:__ The C-statistic is computed using `yardstick::roc_auc`, which wraps `pROC::roc`. To illustrate, if a variable importance metric assigned higher importance for all of the $x$ variables compared to all the $z$ variables, it would have a C-statistic of 1 for the $x$ variables.

## Results

Table 1: Discrimination between relevant and irrelevant predictor variables for each VI technique in 6 different simulated scenarios. Cells that are highlighted with a light blue color indicate which variable importance technique had the best discrimination for a given sample size, correlation setting, and variable type. For example, ANOVA importance had the best discrimination for linear effects with N = 2,500 and zero correlation between the linear effects and irrelevant variables. However, when the correlation between linear effects and irrelevant variables was non-zero, negation importance had the best discrimination across all scenarios and for each effect type. Given the low likelihood of ever having a real dataset with zero correlation between the predictor variables, these data suggest that negation importance can be very beneficial in applied settings.

data_gt <-  
  benchmark_vi_smry |>
  ungroup() |> 
  mutate(xcorr = recode(xcorr, 
                        '0.0' = '|Corr(linear effects, irrelevant effect)| = 0',
                        '0.1' = '|Corr(linear effects, irrelevant effect)| \u2264 0.10',
                        '0.2' = '|Corr(linear effects, irrelevant effect)| \u2264 0.20',
                        '0.3' = '|Corr(linear effects, irrelevant effect)| \u2264 0.30',
                        '0.4' = '|Corr(linear effects, irrelevant effect)| \u2264 0.40'),
         model = recode(model,
                        'aorsf' = 'Negation importance',
                        'aorsf_menze' = 'ANOVA importance',
                        'xgboost' = 'SHAP importance',
                        'randomForestSRC' = 'Permutation importance')) |> 
  select(n_obs, model, xcorr, overall, x, w, g, v) |> 
  pivot_wider(names_from = n_obs, 
              values_from = c(overall, x, w, g, v)) |> 
  select(xcorr,
         model,
         overall_2500,
         x_2500,
         w_2500,
         g_2500,
         v_2500,
         overall_5000,
         x_5000,
         w_5000,
         g_5000,
         v_5000) |> 
  arrange(xcorr, model)

cols_to_highlight <- c("overall_2500",
                       "x_2500",
                       "w_2500",
                       "g_2500",
                       "v_2500",
                       "overall_5000",
                       "x_5000",
                       "w_5000",
                       "g_5000",
                       "v_5000")

max_row_positions <- data_gt |>
  group_by(xcorr) |> 
  summarize(across(all_of(cols_to_highlight), which.max)) |> 
  mutate(xcorr = c(0, 4, 8, 12, 16)) |> 
  pivot_longer(cols = -xcorr) |> 
  mutate(value = value + xcorr) |> 
  select(name, value)

code_styles <- map2_chr(
  .x = max_row_positions$value,
  .y = max_row_positions$name,
  .f = ~ paste0(
    "|>",
    "tab_style(style = cell_fill(color = 'lightblue')",
    ", locations = cells_body(columns = ", .y, ",rows = ", .x, "))"
  )
)

results_tbl_1 <- data_gt |> 
  mutate(across(ends_with("0"), ~table_value(100*.x))) |> 
  gt(rowname_col = 'model', groupname_col = 'xcorr') |> 
  tab_stubhead(label = 'Variable importance technique') |> 
  tab_spanner(label = 'N = 2,500', 
              columns = c("overall_2500",
                          "x_2500",
                          "w_2500",
                          "g_2500",
                          "v_2500")) |> 
  tab_spanner(label = 'N = 5,000', 
              columns = c("overall_5000",
                          "x_5000",
                          "w_5000",
                          "g_5000",
                          "v_5000")) |> 
  cols_label(
    overall_2500 = "All effects",
    x_2500 = "Linear effects",
    w_2500 = "Non-linear effects",
    g_2500 = "Interaction effects",
    v_2500 = "Combo effects",
    overall_5000 = "All effects",
    x_5000 = "Linear effects",
    w_5000 = "Non-linear effects",
    g_5000 = "Interaction effects",
    v_5000 = "Combo effects"
  ) |> 
  cols_align(align = 'center')

code_to_run <- paste0('results_tbl_1', 
                      paste0(code_styles, collapse = ''))

eval(rlang::parse_expr(code_to_run))

### Overall summary

Due to space constraints, the pilot application presents a summary of the results in Table 1. This summary aggregates the results from the 6 scenarios presented above, and it is shown here here to verify.

Table 2: Discrimination between relevant and irrelevant predictor variables for each VI technique. Negation importance obtains the highest C-statistic for each variable type except non-linear variables, where it closely trails SHAP importance.

data_fig <- benchmark_vi_smry |>
  ungroup() |> 
  mutate(model = recode(model,
                        'aorsf' = 'Negation importance',
                        'aorsf_menze' = 'ANOVA importance',
                        'xgboost' = 'SHAP importance',
                        'randomForestSRC' = 'Permutation importance')) |> 
  select(n_obs, model, xcorr, overall) |> 
  group_by(n_obs, model, xcorr) |> 
  summarize(value = mean(overall))

ggplot(data_fig) + 
  aes(x = xcorr, y = value, col = model) + 
  geom_line() +
  facet_wrap(~n_obs)

results_tbl_2 <- benchmark_vi_smry |>
  ungroup() |> 
  mutate(model = recode(model,
                        'aorsf' = 'Negation importance',
                        'aorsf_menze' = 'ANOVA importance',
                        'xgboost' = 'SHAP importance',
                        'randomForestSRC' = 'Permutation importance')) |> 
  select(n_obs, model, xcorr, overall, x, w, g, v) |> 
  group_by(model) |> 
  summarize(
    across(
      c(overall, x, w, g, v),
      ~table_value(mean(.x))
    )
  ) |> 
  gt(rowname_col = 'model') |> 
  tab_stubhead(label = 'Variable importance technique') |> 
  cols_label(
    overall = "Overall",
    x = "Linear effects",
    w = "Non-linear effects",
    g = "Interaction effects",
    v = "Combo effects"
  ) |> 
  cols_align(align = 'center')

results_tbl_2

@@

The strengths of negation importance are:

1. It can be computed for any type of oblique random forest, as it only leverages the linear coefficients in the nodes. (ANOVA importance requires a p-value be computed for each variable in each node).

1. It penalizes over-fitting by evaluating performance in out-of-bag data.

1. In the current simulation, it had the most reliable estimate of variable importance when irrelevant variables were correlated to relevant ones.
